---
title: "SEEFLEX Cross Validation"
author: "Tobias Pauls"
date: "2025-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#### Set working directory to the SEEFLEX root folder ####

```

# Data preparation and packages

```{r cars}

library(MASS)
library(caret)
library(e1071)
library(tidyverse)

# Source utilities files
source("code/gma_analysis/gma_utils.R")
source("code/gma_analysis/seeflex_gma_utils.R")

# Load data file
load("data/gma/20250412_data.rda", verbose=TRUE)

# Filter for texts of min 10 sentences and min 100 words
res2 <- filter_text_length_and_adjust_meta(
  seeflex_features_original, seeflex_meta, 100, 5)
seeflex_features_original <- res2$feat_df
seeflex_meta <- res2$meta_df

# Deselect the word, sentence and token count and create GMA matrix
seeflex <- convert_to_gma_matrix(seeflex_features_original)
seeflex_z <- scale(seeflex)
seeflex_zl <- signed.log(seeflex_z)

# Perform LDA on data
PCA <- GMA(seeflex_zl)
LDA <- PCA$copy()
LDA$add.discriminant(seeflex_meta$GENRE)
seeflex_CV <- LDA$projection("both")

# Convert dataset to df
seeflex_CV_lda <- as.matrix(seeflex_CV[,1:4])
seeflex_CV_df <- as.data.frame(seeflex_CV[, 1:4])

# Define a response variable and convert object class to factor
response <- seeflex_meta$GENRE
if (!is.factor(response)) {
  response <- as.factor(response)
}

# Define number of observations
nrow_seeflex <- nrow(seeflex_CV)

```

# Support Vector Machine

```{r pressure, echo=FALSE}

res.type <- svm(
  seeflex_CV_lda, 
  response, 
  kernel = "radial", 
  cross = 10)

svm.report <- function (res) {
  acc <- mean(res$accuracies)
  cat(sprintf("Mean accuracy: %.1f%%\n", acc))
  cat("Cross-validation folds:\n")
  print(round(res$accuracies, 1))
}

svm.report(res.type)

```

# Leave-one-out cross validation (LOOCV)

NB: This step is included in the markdown as replication of the cross validation technique applied by Diwersy et al. (2014). This is problematic with the SEEFLEX data, however, since the data cannot be assumed to be independent due to various samples per participant and grouping into e.g. registers and genres, as well as schools and school years. Further below, a grouped cross-validation is performed to counteract the violation of the independence assumption (Winter & Grice, 2021)

```{r}

# Initialize a vector to store predictions and initialize a count to store the
# correct assignment to classes
predictions <- character(nrow_seeflex)
count <- 0
sflx <- as.data.frame(seeflex_zl)

# Perform the LOOCV
for (i in 1:nrow_seeflex) {
  
  # Exclude one observation (i) in each loop
  train_data <- sflx[-i, ]
  
  # Fit the LDA model on training data using the correct response variable 
  lda_model <- MASS::lda(response[-i] ~ ., data = train_data)
  # svm_model <- svm(
  #   x = train_data,
  #   y = response[-i],
  #   kernel = "radial",
  #   cross = 10)
  
  # Predict on the left-out observation
  pred <- predict(lda_model, newdata = sflx[i, , drop = FALSE])
  # pred <- predict(svm_model, newdata = sflx[i, , drop = FALSE])
  print(paste("pred", pred$class))
  
  # Store prediction
  predictions[i] <- pred$class
  # predictions[i] <- pred #svm
  
  # Print actual vs predicted for debugging purposes
  print(paste("resp", response[i]))
  if (response[i] == pred$class) {
    count <- count + 1  # Increment the counter
  }
  cat("Observation", i, "- Actual:", response[i], "Predicted:", pred$class, "\n")
}

# Compare predictions with actual values and calculate accuracy
accuracy <- count / nrow_seeflex * 100
cat("Accuracy of LDA with Leave-One-Out Cross-Validation:", accuracy, "%\n")

```

# Grouped cross validation (GCV)

For the grouped cross validation, the group is included in the data frame and the data is partitioned with knowledge of the grouping variable.

```{r}

seeflex_gcv <- as.data.frame(seeflex_zl)
seeflex_gcv <- seeflex_gcv %>% 
  mutate(group = response)

set.seed(123)
# folds <- groupKFold(seeflex_gcv$group, k = 5)  # You can change 'k' as needed
folds <- createDataPartition(seeflex_gcv[,"group"],p=.75,list=FALSE)

# Initialize results storage
results <- list()

# Perform cross-validation
for (i in seq_along(folds)) {
  train.index <- unlist(folds[-i])  # Get training indices for this fold
  
  if (length(unique(seeflex_gcv$group[train.index])) < length(levels(seeflex_gcv$group))) {
    warning(paste("Fold", i, "is missing some group levels"))
    next  # Skip this iteration if necessary
  }
  
  seeflex_gcv.trn <- seeflex_gcv[train.index, ]
  seeflex_gcv.tst <- seeflex_gcv[folds[[i]], ]  # Test set for this fold

  ctrl <- trainControl(method = "none")  # No internal resampling needed here
  
  fit.cv <- train(group ~ ., data = seeflex_gcv.trn, method = "lda",
                  trControl = ctrl) 

  pred <- predict(fit.cv, seeflex_gcv.tst)
  
  cm <- confusionMatrix(table(seeflex_gcv.tst[,"group"], pred))
  
  results[[i]] <- cm$overall['Accuracy']  # Store accuracy or any other metric you're interested in
}

# Print results of each fold
print(results)

# If you want an overall accuracy across all folds:
mean_accuracy <- mean(unlist(results))
cat("Mean Accuracy across all folds:", mean_accuracy, "\n")

```

## References

Diwersy, S., Evert, S., & Neumann, S. (2014). A weakly supervised multivariate approach to the study of language variation. In B. Szmrecsanyi & B. Wälchli (Hrsg.), Aggregating Dialectology, Typology, and Register Analysis (S. 174–204). DE GRUYTER. https://doi.org/10.1515/9783110317558.174.

Winter, B., & Grice, M. (2021). Independence and generalizability in linguistics. Linguistics, 59(5), 1251–1277. https://doi.org/10.1515/ling-2019-0049.

